decompr: GVC decomposition in R

### 2015-01-29 draft.Rnw
update text

\documentclass[a4paper,twoside]{article}

@@

\vspace{\stretch{1}}
\abstract{Global Value Chains have become a central unit of analysis in research on international trade.
However, the complex matrix transformations at the basis of most Value Chain indicators still constitute a significant entry barrier to
the field.
The R package \textbf{decompr} solves this problem by implementing algorithms for the analysis of Global Value Chains,
which calculate standard indicators in an automated fashion. Two methods for gross export flow decomposition using 
Inter-Country Input-Output tables are provided.
The first method is the \textbf{Wang-Wei-Zhu} algorithm which splits bilateral gross exports into 16 value added components. These components can broadly be divided into domestic and foreign value added in exports.
The second method concerns a decomposition based on the classical \citet{wale36} insight.
It derives the value added origins of an industry's exports by source country and source industry, using easily available
gross trade data.
This article summarises the methodology of the algorithms, describes the format of the input and output data, 
and exemplifies the usefulness of the two methods on the basis of a simple example data set.}
\pagebreak

\section{Introduction}
The analysis of Global Value Chains (GVCs) has become a central topic in research on trade and development policy.
Both policy makers and academia increasingly value the growth opportunities GVCs offer to world trade and, especially, to developing countries.
However, analysing this phenomenon empirically requires complex matrix manipulations, since the relevant data is only available in the form of gross flows.
This package enables researchers with little background in matrix algebra and linear programming to easily derive standard GVC indicators for statistical analysis.

\textbf{decompr} uses Inter-Country Input-Output tables (ICIOs), such as those published by the OECD, 
the World Input Output Database (WIOD),  or national statistics bureaus, as input. 
These tables state supply and demand relationships in gross terms between industries within and across countries.

@@

The \textbf{Wang-Wei-Zhu} (henceforth WWZ) decomposition goes a step further by not only revealing the source of the value added,
but also breaking down exports into different categories, based on its final usage and destination.
It implements the theoretical work of \citet{wang2014quantifying}. The main categories are:

@@

The next section introduces the data as it is used by the package, as well as an example data set provided by the WIOD.
Section three summarises the theoretical derivation for the decompositions, and shows how these can be performed in R using the package.
We conclude with a discussion of potential uses and limitations of this approach.

\section{Data}
\label{sec:data}
Two data sets are included in the package, one real world data set, and one minimal fictional data, for training surposes.
The forme is the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
The later, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify the mechanics of the package.

@@

As can be seen above, a \textbf{decompr} class object is in fact a list containing thirty different objects. Depending on the choice of decomposition, all or some of these objects are used.

\section{Leontief decomposition}

Let us now turn to the actual algorithms starting with the \textbf{Leontief} decomposition. We shortly describe the theoretical
derivation of the method to expose the inner steps of the \textbf{decompr} package. Then we turn to the technical implementation
and finish with a description of the output.

The tools to derive the source decomposition date back to \citet{wale36} who showed that, with a set of simple calculations, 
national Input-Output tables based on gross terms give the true value added flows between industries. 
The idea behind this insight is that the production of industry \textbf{i}'s output requires inputs of other industries 
and \textbf{i}'s own value added. The latter is the direct contribution of \textbf{i}'s output to domestic value added. 
The former refers to the first round of \textbf{i}'s indirect contribution to domestic value added since the input from 
other industries that \textbf{i} requires for its own production triggers the creation of value added in the supplying 
industries. As the supplying industries usually depend on inputs from other industries, this sets in motion a second round 
of indirect value added creation in the supplying industries of the suppliers, which is also caused by \textbf{i}'s production. 
This goes on until value added is traced back to the original suppliers and can mathematically be expressed as

@@

\section{Wang-Wei-Zhu decomposition}
The Wang-Wei-Zhu decomposition ...

@@

\section{Conclusion}

@@

\section{Tables}

>>>>

\documentclass[a4paper]{article}

@@

\vspace{\stretch{1}}
\abstract{Global Value Chains have become a central unit of analysis in research on international trade.
However, the complex matrix transformations at the basis of most Value Chain indicators still constitute a significant entry barrier to
the field.
The R package \textbf{decompr} solves this problem by implementing the algorithms for the analysis of Global Value Chains as R procedures, thereby simplifying the decomposition process..
Two methods for gross export flow decomposition using Inter-Country Input-Output tables are provided.
The first method is the \textbf{Wang-Wei-Zhu} algorithm which splits bilateral gross exports into 16 value added components. These components can broadly be divided into domestic and foreign value added in exports.
The second method concerns a decomposition based on the classical \citet{wale36} insight.
It derives the value added origins of an industry's exports by source country and source industry, using easily available
gross trade data.
This article summarises the methodology of the algorithms, describes the format of the input and output data, 
and exemplifies the usefulness of the two methods on the basis of a simple example data set.}
\pagebreak

\section{Introduction}
Global Value Chains (GVCs) have become a central topic in research on trade and development policy.
Both policy makers and academia increasingly value the growth opportunities GVCs offer to global trade and, especially, to developing countries.
However, analysing this phenomenon empirically requires complex matrix manipulations, since the relevant data is only available in the form of gross flows.
This package enables researchers with little background in matrix algebra and linear programming to easily derive standard GVC indicators for statistical analysis.

The \textbf{decompr} package uses Inter-Country Input-Output tables (ICIOs), such as those published by the OECD and WTO (TiVA), % CITE!! 
the World Input Output Database \parencite{timmer2012world},  or national statistics bureaus, as input.
These tables state supply and demand relationships in gross terms between industries within and across countries.

@@

The \textbf{Wang-Wei-Zhu} (henceforth WWZ) decomposition goes a step further by not only revealing the source of the value added, but also breaking down exports into different categories, according to final usage and destination.
It implements the theoretical work of \textcite{wang2014quantifying}.
The high-level divisions in this framework are listed below.

@@

\Cref{sec:data} introduces the data as it is used by the package, as well as an example data set provided by the WIOD.
After which \cref{sec:leontief,sec:wwz} summarise the theoretical derivation for the two decompositions, and show how these can be performed in R using \textbf{decompr}.
We conclude with a discussion of potential uses and limitations of this approach.

\section{Data}
\label{sec:data}
Two data sets are included in the package, one real world data set, and one minimal data set (fictional), for training surposes.
The former is the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
The later, is a fictional 3-country 3-sector data set, which we will use throughout this article to demonstrate the usage of the decompr package.

@@

As can be seen above, a \textbf{decompr} class object is in fact a list containing thirty different objects.
Depending on the choice of decomposition, all or some of these objects are used

\section{Leontief decomposition}
\label{sec:leontief}
Let us now turn to the actual algorithms starting with the \textbf{Leontief} decomposition. We shortly describe the theoretical
derivation of the method to expose the inner steps of the \textbf{decompr} package. Then we turn to the technical implementation
and finish with a description of the output.

\subsection{Theoretical derivation}
The tools to derive the source decomposition date back to \citet{wale36} who showed that, with a set of simple calculations, 
national Input-Output tables based on gross terms give the true value added flows between industries. 
The idea behind this insight is that the production of industry \textbf{i}'s output requires inputs of other industries 
and \textbf{i}'s own value added. The latter is the direct contribution of \textbf{i}'s output to domestic value added. 
The former refers to the first round of \textbf{i}'s indirect contribution to domestic value added since the input from 
other industries that \textbf{i} requires for its own production triggers the creation of value added in the supplying 
industries. As the supplying industries usually depend on inputs from other industries, this sets in motion a second round 
of indirect value added creation in the supplying industries of the suppliers, which is also caused by \textbf{i}'s production. 
This goes on until value added is traced back to the original suppliers and can mathematically be expressed as

@@

\section{Wang-Wei-Zhu decomposition}
\label{sec:wwz}
The Wang-Wei-Zhu decomposition ...

@@

\section{Conclusion}
\label{sec:conclusion}

@@

\section{Tables}
\label{sec:tables}

### 2015-01-29 draft.tex
line breaks by sentence
update text

\documentclass[a4paper,twoside]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

The next section introduces the data as it is used by the package,
as well as an example data set provided by the WIOD.
Section three summarises the theoretical derivation for the decompositions,
and shows how these can be performed in R using the package.
After which we conclude with a discussion of potential uses and limitations of this approach.

\section{Data}
\label{sec:data}
Two data sets are included in the package, one real world data set, and one minimal fictional data, for training surposes.
The forme is the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
The later, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify
the mechanics of the package.

>>>>

\documentclass[a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\Cref{sec:data} introduces the data as it is used by the package, as well as an example data set provided by the WIOD.
After which \cref{sec:leontief,sec:wwz} summarise the theoretical derivation for the two decompositions, and show how these can be performed in R using \textbf{decompr}.
We conclude with a discussion of potential uses and limitations of this approach.

\section{Data}
\label{sec:data}
Two data sets are included in the package, one real world data set, and one minimal data set (fictional), for training surposes.
The former is the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
The later, is a fictional 3-country 3-sector data set, which we will use throughout this article to demonstrate the usage of the decompr package.
